---
title: "Create Performance Table for your Prediction Model"
description: |
  How to create a Performance Table from scratch
author:
  - name: Uriah Finkel
    url: https://example.com/norajones
date: 2021-11-10
output:
  distill::distill_article:
    md_extensions: +emoji
    toc: true
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ISLR)
library(gt)
library(rsample)
library(glue)
```

In the upcoming {rtichoke} Package there is a function that is dedicated for creation of a performance table from a combination of predictions and outcomes:

`rtichoke::create_performance_table()`.

## Motivation

I had several motivations for writing this post:

### :bulb:Introduce some theoretical insights:

There are a lot of great resources on data science, statistics and performance metrics online - but I think that many of the insightful discussions are not so easy to find. In this tutorial I'll try to mention some of the related key concepts without going into details, in order for the reader to understand how these concepts take place in the big picture. TODO: rephrase.

### :package: Introducing some of {gt} capabilities

Now that our performance data is ready, we can render a table with the {gt} package.\
There are many great packages for table creation available for r, but I think that {gt} is the most convenient one because I'm so used to the "grammer of graphics" from {ggplot2}. {gt} works in a similar way, you start from basic table and you can add all kinds of stuff along the way. If you are a {tidyverse} user {gt} should be very convenient to you two!\

--In the following post I'll show the different stages . -- TODO: add motivation for the post

## Prepare Performance Data :dart:

### Chop probabilities to binary predictions :knife:

A basic problem in delivering performance metrics for predictions of binary outcome is that we are estimating a probability, which is in fact a continuous number **between 0 and 1** ( $0 \leq\hat{p} \leq 1$) while the outcome **must be 0 or 1** ( $y∈\{1, 0\}$ ).

```{r echo=TRUE, message=FALSE, warning=FALSE}

predictions_and_outcomes <- tibble(probs = c(0.21, 0.45, 0.13, 0.02, 0.34),
       real =  c(0, 1, 0, 1, 0)) 

predictions_and_outcomes %>% 
  gt()

```

Most of Performance Metrics for probability estimation demand a threshold that will make the prediction also a binary. Generally speaking it's easier to grasp True or False in terms of predictions, even though it might come with a price (TODO: add link). Some performance metrics such as the Brier-Score (TODO: add link) or the C-Index (more familiar as the Area under the ROC curve, AUC or AUROC) (TODO: add link) do not demand dichotomy, but they might not be as intuitive as performance metrics that are based on dichotomisation of the predictions.

Let's say that we chose a probability threshold of 0.15. In that case when the estimated probability is higher than 0.15 we will classify the prediction as *Positive* and if it's less or equal to 0.15 we will classify it as *Negative*:

```{r echo=TRUE, message=FALSE, warning=FALSE}

predictions_and_outcomes <- predictions_and_outcomes %>% 
  mutate(binary_prediction = as.numeric(probs > 0.15))

predictions_and_outcomes %>% 
  gt()

```

### Add Confusion Matrix :confused:

Now we can classify each prediction to the appropriate case in what is known as the *Confusion Matrix*:

```{r}
library(gt)
library(tidyverse)

tribble(
  ~" ", ~"Predicted Positive", ~"Predicted Negative",
  "Real Positive", "TP", "FN",
  "Real Negative", "FP", "TN"
) %>%
  gt::gt(rowname_col = " ") %>%
  cols_align(
    align = c("center"),
    columns = c("Predicted Positive", "Predicted Negative")
  ) %>%
  tab_style(
    style = cell_fill(color = "pink"),
    locations = list(
      cells_body(
        columns = "Predicted Positive",
        rows = "Real Negative"
      ),
      cells_body(
        columns = "Predicted Negative",
        rows = "Real Positive"
      )
    )
  ) %>%
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = list(
      cells_body(
        columns = "Predicted Positive",
        rows = "Real Positive"
      ),
      cells_body(
        columns = "Predicted Negative",
        rows = "Real Negative"
      )
    )
  )
```

**TP** (True Positive) - when the prediction is 1 (*positive*) and the outcome is indeed 1.

**TN** (True Negative) - when the prediction is 0 (*negative*) and the outcome is indeed 0.

**FP** (False Positive) - when the prediction is 1 but the outcome is 0.

**FN** (False Negative) - when the prediction is 0 but the outcome is 1.

```{r echo=TRUE, message=FALSE, warning=FALSE}

predictions_and_outcomes <- predictions_and_outcomes %>% 
  mutate("Confusion Matrix" = case_when(
    real == 1 & binary_prediction == 1 ~ "TP",
    real == 0 & binary_prediction == 0 ~ "TN",
    real == 1 & binary_prediction == 0 ~ "FN",
    real == 0 & binary_prediction == 1 ~ "FP"
  ))

predictions_and_outcomes %>% 
  gt()

```

So far it all seems pretty straight forward, but usually we would like to see accuracy among many different Probability Thresholds. For our example we will use the `Carseats` data set from the `ISLR` package that provides a collection of data sets to the highly recommended [ISLR book](https://www.statlearning.com/).

For this tutorial I will use the code from [ISLR tidymodels Labs](https://emilhvitfeldt.github.io/ISLR-tidymodels-labs/index.html) (which is also highly recommended). We will try to predict high sales by dichotomising the `Sales` variable, and split the data to train and test set with the help of the `rsample` package:

```{r}

library(ISLR)
data(Carseats)

Carseats <- as_tibble(Carseats) %>%
  mutate(High = as.numeric(Sales <= 8)) %>%
  select(-Sales)

set.seed(1234)
Carseats_split <- initial_split(Carseats)
Carseats_train <- training(Carseats_split)
Carseats_test <- testing(Carseats_split)

```

We are ready to use our first model:

### Logistic Regression :twisted_rightwards_arrows:

Logistic Regression is the classic solution for prediction of binary outcome, it provides an explainable equation and [it is used for many years](https://papers.tinbergen.nl/02119.pdf).

I'm a big fan of `tidymodels` because it provides a solution to the exhausting [syntactical diversity](https://www.tmwr.org/base-r.html#tidiness-modeling) for models in R, but sometimes it's an overkill for GLM models - so for this part I'll use some classic R code:

```{r}

logistic_regression_model <- glm(High ~ ., 
                                 data = Carseats, 
                                 family = "binomial")

predictions_and_outcomes <- tibble(probs = predict(logistic_regression_model, 
                                           type = "response"), 
                           real = Carseats$High)


predictions_and_outcomes %>% 
  head() %>% 
  gt()
```

TODO: maybe a better code? PURRR?

We are ready to build our Performance Data, For each Probability Threshold we will calculate the confusion matrix:

```{r}

N <- nrow(predictions_and_outcomes)
real_positives <- sum(predictions_and_outcomes$real)
real_negatives <- N - real_positives
prevalence <- real_positives / N

performance_data <- tibble(
  threshold = seq(0, 1, by = 0.1)
) %>% 
  mutate(
    TP = lapply(threshold, function(x) sum(predictions_and_outcomes$probs[predictions_and_outcomes$real == 1] > x)) %>%
        unlist(),
    TN = lapply(threshold, function(x) sum(predictions_and_outcomes$probs[predictions_and_outcomes$real == 0] <= x)) %>%
      unlist(),
    FN = real_positives - TP,
    FP = real_negatives - TN
  )

performance_data %>% 
  head() %>%
  gt()

```

**Stratified performance by Probability Threshold or by Predicted Positives Condition Rate?**

Sometimes you have a good reason to decide on a specific Probability Threshold, but sometimes you have some resource in scarce. Let's say that you have a specific constant amount of medicine that you can't give to everyone, but you want to identify the populations that are at the most risk.

In that case you will not care much about the specific threshold, you would rather to see performance by The Predicted Positive Condition Rate (PPCR), which is the fraction of Predicted Positives if you were to pick them by their highest probability to have an event.

### Add Performance Metrics :straight_ruler:

Now that we have a confusion matrix for each Probability Threshold we can calculate some common performance metrics.

-   **Sensitivity** and **Specificity** are extremely popular, but you should notice that in terms of prediction they are [conditional on the future.](https://www.fharrell.com/post/backwards-probs/)

$$
\text{Sensitivity} = \frac{\text{TP}}{\text{TP + FN}} = \frac{\text{TP}}{\text{Real Positives}} = \text{Prob( Predicted Positive | Real Positive )}
$$

$$
\text{Specificity} = \frac{\text{TN}}{\text{TN + FP}} = \frac{\text{TN}}{\text{Real Negatives}} = \text{Prob( Predicted Positive | Real Positive )}
$$ A variation of Specificity is called the False Positive Rate (FPR) and it is nothing more but \$\$1 - Specificity\$\$. It's is used in the ROC curve, but honestly I don't think that it's very useful. The thing that you do want to remember is that unlike Sensitivity and Specificity you want FPR to be as small as possible.

-   **PPV** and **NPV** might be more suitable for prediction model because they are conditional on your predictions (and you can decide *now* how much observations you want to predict as positives):

    $$
    \text{PPV} = \frac{\text{TP}}{\text{TP + FP}} = \frac{\text{TP}}{\text{Predicted Positives}} = \text{Prob( Real Positive | Predicted Positive )}
    $$

    $$
    \text{NPV} = \frac{\text{TN}}{\text{TN + FN}} = \frac{\text{TN}}{\text{Predicted Negatives}} = \text{Prob( Real Negative | Predicted Negative )}
    $$

-   **Prevalence** is *not* a performance metric, but it is a key concept in order to understand how much the data is balanced. It is in fact nothing more but the fraction of positive observations from the whole population and therefore it is constant regardless the probability threshold.

    $${\text{Prevalence}} = \frac{\text{TP + FN}}{\text{TP + FP + TN + FN}}$$

-   **Lift** is the ratio between **PPV** and the **Prevalence**, or in other words - how much you model is better than a random guess?

    $$
    \text{Lift} = \frac{\text{PPV}}{\text{Prevalence}} = \frac{\cfrac{\text{TP}}{\text{TP + FP}}}{\cfrac{\text{TP + FN}}{\text{TP + FP + TN + FN}}} 
    $$

-   **Net-Benifit** is (by my view) an underused metric. I think it's a powerful way to derive Utility in a generic way. There is an assumption that says that your probability threshold in fact includes the exchange right between False Positives (FP) and False Negatives (FN). TODO: add formula and add reference.

    \
    :warning: **Warning: You don't want to use Net Benefit based on Predicted Positives Condition Rate!** **If you are using Net Benefit you must declare the explicit Probability Threshold, because it is part of the components of the metric!**\
    \
    Another Unique think about Net-Benefit is that it can be Negative or Positive. Negative means that you'd rather not use the model at all, and Positive means that you should use the model.

    Also keep in mind that this performance metric also shows if your model is well-calibrated: You'll see that [when we will compare different models](#compare-models-⚖️) how Net-Benefit varies.

    Any way I recommend on reading more about this performance metric.

```{r}

performance_table <- performance_data %>%
    dplyr::mutate(
      sensitivity = TP / (TP + FN),
      FPR = FP / (FP + TN),
      lift = (TP / (TP + FN)) / ((TP + FP) / N),
      specificity = TN / (TN + FP),
      PPV = TP / (TP + FP),
      NPV = TN / (TN + FN),
      predicted_positives = TP + FP,
      PPCR = predicted_positives / N,
      NB = TP / N - (FP / N) * (threshold / (1 - threshold))
    ) %>% 
  gt()
  
performance_table 

```

## Style the performance table with {gt} :package: {gt=""}

Now that we have a performance table we can stylize it: Tables should be accessible to the reader, some of the things that are useful to the analyst are not useful at all to the reader.

The philosophy behind the syntax in {gt} (*grammar of tables)* is similar to the one behind {ggplot2} (*grammar of graphics)*, we are starting from a basic table and adding another layer of code each time.

**Format Missing Values: `fmt_missing()`**

Since `NaN` is not very pleasing in tables, we can change it to empty string.

But why some of the values of Lift, PPV, NPV and Net Benefit are missing?

-   Lift and PPV are conditional on Positive Prediction, if there Probability Threshold is higher than all of the probabilities there are no Positive Predictions and nothing to condition on.

-   NPV is conditional on Negative Prediction, if the Probability Threshold is lower than all of the probabilities there are no Negative Predictions and nothing to condition on.

-   Net Benifit assumes that the odds of the Probability Threshold stands for exchange rate between TP and TN, when the Probability Threshold is 1 the cost of TN is infinite and Net Benfit is not defined.

**Format 2 decimals for each metric: `fmt_number()`**

For performance metric 2 decimals are probably more than enough.

**Format values as a percentage and merge with counts: `cols_merge_n_pct()`**

There is no need for two different columns for percentage and counts of the predicted positives.

**Relabel Columns: `cols_label()`**

It might differ from one user to another, but when I write columns names in a data frame I use [snake_case](https://twitter.com/allison_horst/status/1205702878544875521), and when I display column names in a table I use capital letters.

**Align Columns: `cols_align()`**

There are different thoughts about alignment, but the most important thing is to be aware to it and ensure consistency.

**Adding spanner column labels: `tab_spanner()`**

Tab spanner is a good way to separate different kinds of columns, performance metrics are derived from the confusion matrix therefore it make sense to set a different spanners.

**Set different background color: `tab_options()`**

White is the default choice for background, but in my opinion it is overused.

**Set title and subtitle: `tab_header()`**

Without a lot of effort the title can be styled with markdown syntax (\* for *italics*, \*\* for **bold**).

```{r echo=TRUE}

performance_table <- performance_table %>%
    fmt_missing(
      columns = c(lift, PPV, NPV, NB),
      missing_text = ""
    ) %>%
    fmt_number(columns = c(sensitivity, FPR, lift, specificity, 
                           PPV, NPV, PPCR, NB),
               decimals = 2) %>%
  fmt_percent(
    columns = PPCR
  ) %>% 
  cols_merge_n_pct(
    col_n = predicted_positives,
    col_pct = PPCR
  ) %>%
    cols_label(
      sensitivity = "Sens",
      lift = "Lift",
      specificity = "Spec",
      predicted_positives = "Predicted Positives",
      FPR = "1 - Spec"
    ) %>%
    cols_align(
      align = "left",
      columns = dplyr::everything()
    ) %>% 
    tab_spanner(
      label = "Confusion Matrix",
      columns = c(
        TP, FP, TN, FN)
    ) %>%
    tab_spanner(
      label = "Performance Metrics",
      columns = c(
        sensitivity, lift, specificity, FPR,
        PPV, NPV, NB
      )
    ) %>%
    tab_options(
      table.background.color = "#FFFBF3"
    ) %>%
    tab_header(
      title = gt::md("**Performanc Metrics for Different Thresholds**"),
      subtitle = gt::md("*Logistic Regression*")
    )

performance_table
```

## Compare Models :balance_scale:

So far we examined the performance of a **Logistic Regression :twisted_rightwards_arrows:** for our population. We will explore the performance of two additional models:

### Tree Model :evergreen_tree:

Tree Models are extremely popular in Machine Learning, they are considered to be "Black Boxes" because as a user you don't really know (or even care) what they do, but you know that they are doing it well in terms of accuracy.

```{r}

```

### Random Guess :see_no_evil:

Emile Borel once claimed:

> If an army of monkeys were strumming on typewriters they might write all the books in the British Museum.

We don't have an army of monkeys, but for education purposes we will give one monkey to use the `rbinom()` function. In order to be fair we will tell the monkey the prevalence of the population and set it as the *p* parameter in the binomial distribution.

## Add Colors to Title and Subtitle :art:

## Background Bars

## Several Populations 👩👨

There are many reasons to explore results for different populations: The most common in Machine Learning models is when the analyst want to avoid overfitting the model. Logistic Regression will probably will not overfit the data anyway, but decision-tree models might.

Another possible case is that when you want to assure Fairness in medicine. You'll might have some great results on specific populations that are .

Luckily, we saved some data aside exactly that.
